\selectlanguage{english}
\chapter[C\# analyzer module]{C\# analyzer module (\texttt{analyzer.ts})}
\label{chap3:analyzer}
\begin{chapquote}{Alfred V. Aho}
``A parser generator is a tool for constructing compilers; the syntax tree is the central representation.''
\end{chapquote}

\section{Purpose and responsibilities}
\label{chap3:purpose}

The analyzer module is a \textbf{pure, VS Code--agnostic} TypeScript library located at \texttt{src/csharp/analyzer.ts}. Its contract is narrow: given a UTF-8/UTF-16 C\# source string, return a \texttt{CSharpDiAnalysis} value. No file I/O, no workspace knowledge, no knowledge of \texttt{Microsoft.Extensions.DependencyInjection} beyond the shape of \texttt{AddSingleton|Scoped|Transient} generic calls present \emph{in the same file}.

Responsibilities include:
\begin{itemize}
	\item discovering all instance constructors on \texttt{class\_declaration} nodes in the parse tree;
	\item recording each parameter's textual type name and source span for diagnostic underlining;
	\item emitting \textbf{ConcreteTypeIssue} when a parameter type names a class defined in-file but not an interface defined in-file;
	\item emitting \textbf{CircularDependencyIssue} when in-file classes depend on each other through constructor parameter types in a directed cycle;
	\item emitting \textbf{MissingRegistrationIssue} when a parameter type is not found among \texttt{Add*} generic arguments scanned in the same file;
	\item enforcing the \textbf{32\,KiB parse guard} and try/catch around native parse calls so callers always receive a structured result.
\end{itemize}

The extension host invokes \texttt{analyzeCSharp()} from three contexts: per-document diagnostics on edit, the ``analyze current file'' command, and per-file iteration inside \texttt{buildWorkspaceSummary()}. Workspace-level missing-registration logic \textbf{duplicates and extends} the analyzer's registration concept in \texttt{extension.ts}; the analyzer's missing-registration issue remains strictly \emph{per-file} by design.

\section{Key data structures and interfaces}
\label{chap3:data-structures}

Table~\ref{tab:analyzer-types} lists exported interfaces. Each type is a plain TypeScript interface without classes, favoring JSON-serializable snapshots for potential future testing.

\begin{table}[htbp]
\centering
\caption{Exported interfaces in \texttt{analyzer.ts}}
\label{tab:analyzer-types}
\begin{tabular}{|l|p{8.5cm}|}
\hline
\textbf{Type} & \textbf{Meaning} \\
\hline
\texttt{ConstructorParam} & Parameter \texttt{type}, \texttt{name}, \texttt{startIndex}, \texttt{endIndex} (tree-sitter byte indices into source). \\
\hline
\texttt{ConstructorInfo} & \texttt{className}, \texttt{params[]}, constructor \texttt{startIndex}/\texttt{endIndex}. \\
\hline
\texttt{ConcreteTypeIssue} & \texttt{className}, \texttt{paramType}, parameter span, message. \\
\hline
\texttt{CircularDependencyIssue} & \texttt{cycle: string[]} class names forming a directed cycle. \\
\hline
\texttt{MissingRegistrationIssue} & \texttt{className}, \texttt{paramType} not in file-local registration set. \\
\hline
\texttt{CSharpDiAnalysis} & \texttt{constructors}, \texttt{errors[]}, three issue arrays. \\
\hline
\end{tabular}
\end{table}

\subsection{Parser singleton}

The module initializes tree-sitter once:

\begin{lstlisting}[caption={Parser initialization (\texttt{analyzer.ts})}]
const parser = new Parser();
parser.setLanguage(CSharp);
\end{lstlisting}

Reusing the parser avoids reloading the grammar per call. Parse trees are local variables discarded after \texttt{analyzeCSharp} returns.

\section{tree-sitter API surface used by the analyzer}
\label{chap3:treesitter-api}

The npm package \texttt{tree-sitter} exposes a small imperative API around an incremental GLR-style parser~\cite{tree-sitter}. The DI Plugin uses only the subset required to walk C\# \texttt{class\_declaration} and \texttt{constructor\_declaration} nodes. Table~\ref{tab:treesitter-api} maps each API element to its role in \texttt{analyzer.ts}.

\begin{table}[htbp]
\centering
\caption{tree-sitter APIs consumed by \texttt{analyzer.ts}}
\label{tab:treesitter-api}
\small
\begin{tabular}{|p{3.6cm}|p{9.8cm}|}
\hline
\textbf{API} & \textbf{Purpose in DI Plugin} \\
\hline
\texttt{new Parser()} & Module-singleton parser instance reused for every \texttt{analyzeCSharp} call. \\
\hline
\texttt{parser.setLanguage(CSharp)} & Attaches the \texttt{tree-sitter-c-sharp} grammar compiled to a \texttt{Language} object; required before \texttt{parse}. \\
\hline
\texttt{parser.parse(source)} & Builds a \texttt{Tree} from the full source string; returns \texttt{Parser.Tree}. Throws or rejects on native errors (caught; 32\,KiB guard prevents \texttt{EINVAL}). \\
\hline
\texttt{tree.rootNode} & Entry node (\texttt{compilation\_unit} or equivalent); starting point for queries. \\
\hline
\texttt{rootNode.hasError()} & When true, grammar recovered from syntax errors; analyzer appends warning but continues traversal. \\
\hline
\texttt{SyntaxNode.type} & String grammar symbol (e.g.\ \texttt{class\_declaration}, \texttt{parameter}); drives filtering. \\
\hline
\texttt{SyntaxNode.startIndex}, \texttt{endIndex} & Byte offsets into \texttt{source}; copied to \texttt{ConstructorParam} and used by host for \texttt{Range}. \\
\hline
\texttt{SyntaxNode.childCount}, \texttt{child(i)} & Manual tree walk fallback when typed helpers unavailable. \\
\hline
\texttt{SyntaxNode.namedChildCount}, \texttt{namedChild(j)} & Preferred for \texttt{parameter\_list} children to skip punctuation nodes \texttt{(}, \texttt{,}. \\
\hline
\texttt{SyntaxNode.children} & Array used in \texttt{findTypeOfParameter} fallback search. \\
\hline
\texttt{SyntaxNode.childForFieldName(name)} & Grammar-field access: \texttt{"name"}, \texttt{"type"}, \texttt{"parameters"} per tree-sitter field labels in C\# grammar. \\
\hline
\texttt{SyntaxNode.descendantsOfType(type)} & Efficient collection of all nodes matching a type under a subtree (classes, constructors, interfaces). \\
\hline
\texttt{source.slice(start, end)} & Application code (\texttt{getText}) turns node spans into type names and identifiers. \\
\hline
\end{tabular}
\end{table}

\subsection{Grammar package \texttt{tree-sitter-c-sharp}}

Import \texttt{import CSharp from "tree-sitter-c-sharp"} supplies the precompiled grammar. The grammar defines which \texttt{node.type} strings appear in trees. The analyzer hard-codes stable type names such as:
\begin{itemize}
	\item \texttt{class\_declaration}, \texttt{interface\_declaration};
	\item \texttt{constructor\_declaration};
	\item \texttt{parameter\_list}, \texttt{parameter}, \texttt{\_parameter\_array};
	\item leaf types \texttt{identifier}, \texttt{generic\_name}, \texttt{nullable\_type}, \texttt{predefined\_type}, \texttt{array\_type}.
\end{itemize}

If the grammar package updates node names, \texttt{nodeTypeMatches} also checks optional \texttt{grammarType} for forward compatibility.

\subsection{Why \texttt{childForFieldName} matters}

Unlike naive text parsing, tree-sitter grammars assign \textbf{field names} to children (documented in \texttt{grammar.js} of the language repo). Example: a \texttt{parameter} node exposes \texttt{childForFieldName("type")} and \texttt{childForFieldName("name")}. The analyzer prefers fields over positional \texttt{child(i)} because punctuation and modifiers vary.

When the \texttt{type} field is absent (older grammar or unusual syntax), \texttt{findTypeOfParameter} scans children before the \texttt{name} node's \texttt{startIndex} for identifier-like types---a defensive heuristic preserving robustness on edge-case parameters.

\subsection{Why \texttt{descendantsOfType} matters}

Nested classes, regions, and namespaces can bury a \texttt{constructor\_declaration} deep in the tree. Calling \texttt{classNode.descendantsOfType("constructor\_declaration")} returns all constructors declared lexically inside a class body without manual recursion errors. The same API on \texttt{rootNode} collects all \texttt{class\_declaration} nodes for interface/class sets.

A manual \texttt{collectDescendants} fallback exists because some tree-sitter versions expose \texttt{descendantsOfType} only on certain node classes; the dual path prevents empty constructor lists on older bindings.

\subsection{Indices: bytes vs.\ VS Code UTF-16}

tree-sitter reports \texttt{startIndex}/\texttt{endIndex} as \textbf{byte offsets} into the UTF-8 source buffer passed to \texttt{parse}. VS Code document positions use UTF-16 code units. For ASCII-only identifiers in sample projects, byte and UTF-16 columns coincide; internationalized identifiers may require conversion in future host code. The thesis samples intentionally use ASCII type names to avoid mismatch during evaluation.

\subsection{APIs deliberately not used}

The DI Plugin does \textbf{not} use tree-sitter's incremental \texttt{parse(oldTree, edits)} API, query DSL (\texttt{tree-sitter-query}), or WASM build. Every analysis is a full re-parse---acceptable for files under 32\,KiB and simplifies correctness when the user edits registration lines between keystrokes. It also does not use \texttt{Tree.walk()} visitor API; TypeScript loops over \texttt{descendantsOfType} results are sufficient.

Figure~\ref{fig:parse-tree} in Chapter~\ref{chap6:illustrations} should depict a collapsed syntax tree for \texttt{CheckoutCoordinator}'s constructor with labeled node types---either exported from tree-sitter CLI or drawn schematically.

\subsection{Comparison to workspace-level types}

\texttt{ConstructorInfo} in the analyzer lacks \texttt{fileName}; the host attaches path metadata when merging into \texttt{WorkspaceDiSummary}. Issue types in the analyzer are file-local; the host may suppress or reinterpret them when presenting workspace reports (e.g.\ framework types).

\section{Algorithms and analysis logic}
\label{chap3:algorithms}

\subsection{Entry point: \texttt{analyzeCSharp(source)}}

The public function executes in order:
\begin{enumerate}
	\item If \texttt{source.length > MAX\_PARSE\_SOURCE\_LENGTH} (32767), return empty constructors and \texttt{errors: ["Source exceeds tree-sitter limit..."]}.
	\item Invoke \texttt{parser.parse(source)} inside try/catch; on failure, return error message without throwing.
	\item If root \texttt{hasError()}, append syntax warning but continue best-effort traversal.
	\item Build in-file \texttt{classes} and \texttt{interfaces} sets by walking \texttt{class\_declaration} and \texttt{interface\_declaration} nodes.
	\item Extract constructors; run concrete-type, cycle, and missing-registration passes.
	\item Return populated \texttt{CSharpDiAnalysis}.
\end{enumerate}

\subsection{Parse guard and error handling}

The constant \texttt{MAX\_PARSE\_SOURCE\_LENGTH = 32767} exists because the \texttt{node-tree-sitter} binding rejects larger buffers with \texttt{EINVAL} (surfacing in VS Code as ``Invalid argument''). Entity Framework \texttt{DataContextModelSnapshot.cs} in MealPlanner exceeds this threshold by an order of magnitude. Skipping AST analysis for such files is mandatory for workspace stability; regex-based registration scanning in the host still processes their text for \texttt{AddDbContext} if present elsewhere.

Parse exceptions from corrupted or non-C\# content are caught and converted to \texttt{errors[]} strings. The host never wraps \texttt{analyzeCSharp} in an uncaught promise rejection.

\subsection{Constructor extraction algorithm}

\textbf{Input:} parse tree root. \textbf{Output:} \texttt{ConstructorInfo[]}.

Steps:
\begin{enumerate}
	\item Collect nodes of type \texttt{class\_declaration} via \texttt{descendantsOfType}, with manual child traversal fallback if grammar version differs.
	\item For each class, read identifier child as \texttt{className}.
	\item Collect \texttt{constructor\_declaration} descendants.
	\item For each constructor, locate \texttt{parameter\_list} and iterate \texttt{parameter} or \texttt{\_parameter\_array} children.
	\item For each parameter, resolve type text using \texttt{findTypeOfParameter(node)}:
	\begin{itemize}
		\item prefer \texttt{childForFieldName("type")};
		\item else scan children before \texttt{name} field for identifier-like nodes;
		\item concatenate \texttt{predefined\_type}, \texttt{identifier}, \texttt{generic\_name} text.
	\end{itemize}
	\item Store parameter \texttt{name} from \texttt{name} field child; record node start/end indices.
\end{enumerate}

\textbf{Limitations:} primary constructors (C\# 12+) may use different node shapes; \texttt{record} types may share \texttt{class\_declaration} in grammar; \texttt{ref}/\texttt{out}/\texttt{params} parameters may yield incomplete type strings.

\subsection{Concrete-type detection}

For each constructor parameter with type string \(t\):
\begin{itemize}
	\item Normalize \(t\) by stripping nullable \texttt{?} suffix where applicable.
	\item If \(t \in \texttt{classes}\) and \(t \notin \texttt{interfaces}\), emit \texttt{ConcreteTypeIssue}.
\end{itemize}

\textbf{Rationale:} depending on a concrete class in a constructor is a DI smell when an abstraction exists; the quick fix path encourages extracting \texttt{I\{Name\}}.

\textbf{Example (BuilderFactorySample):} \texttt{ConcreteMailerService(SmtpEmailSender smtp)} flags \texttt{SmtpEmailSender} because the class is declared in the same file and is not listed as an interface.

\textbf{File-local scope:} if \texttt{IEmailSender} is defined in another file, the analyzer still flags \texttt{SmtpEmailSender} in the consumer file. Workspace-level suggestions may still propose \texttt{IEmailSender, SmtpEmailSender} registration---the warning educates about local constructor shape.

\subsection{Circular dependency detection}

\textbf{Graph construction:} nodes are class names \(C\) declared in the file. Directed edge \(A \rightarrow B\) exists if class \(A\) has a constructor parameter of type \(B\) and \(B\) is also a class in the file (not interface, not primitive).

\textbf{Cycle finding:} depth-first search with recursion stack; on revisiting a stack node, record cycle path. Cycles are deduplicated by sorting node keys to avoid duplicate diagnostics for the same ring.

\textbf{Interpretation:} this detects structural cycles resolvable only by changing constructors within the same compilation unit---for example \texttt{A(B b)} and \texttt{B(A a)}. It does \textbf{not} detect runtime cycles through interfaces registered only in other files.

\subsection{Missing registration detection (file scope)}

The file is scanned with:
\begin{lstlisting}[caption={Registration regex (file-local)}]
/Add(?:Singleton|Scoped|Transient)\s*<([^>]+)>/g
\end{lstlisting}
Generic argument groups are split on commas; each token is trimmed and added to \texttt{registeredInFile}. For each constructor parameter type \(p\):
\begin{itemize}
	\item skip if \(p\) is in a built-in primitive set (\texttt{string}, \texttt{int}, \texttt{bool}, etc.);
	\item if \(p \notin \texttt{registeredInFile}, emit \texttt{MissingRegistrationIssue}.
\end{itemize}

\textbf{Important distinction:} workspace-level analysis uses \texttt{collectExtendedRegistrationTypes} (\texttt{AddDbContext}, \texttt{AddAutoMapper}, Identity, \texttt{typeof} pairs). The analyzer does \textbf{not} implement those patterns to remain dependency-free and small. A file may show file-local missing issues while workspace analysis considers the type registered---developers should trust workspace commands for ASP.NET extension-method registrations.

\section{Extension commands using the analyzer}
\label{chap3:commands}

The analyzer exposes no VS Code commands. Table~\ref{tab:analyzer-invocation} maps host features to analyzer usage.

\begin{table}[htbp]
\centering
\caption{Host features invoking \texttt{analyzeCSharp}}
\label{tab:analyzer-invocation}
\begin{tabular}{|p{4.2cm}|p{8.8cm}|}
\hline
\textbf{Host feature} & \textbf{Analyzer role} \\
\hline
Continuous diagnostics & Full \texttt{CSharpDiAnalysis} mapped to \texttt{Diagnostic[]} on each C\# document change. \\
\hline
\texttt{di-plugin.analyzeFile} & Pretty-print constructors and issues to Output channel. \\
\hline
\texttt{buildWorkspaceSummary} & Per-file merge of constructors; file-local issues optional for aggregate reporting. \\
\hline
Factory eligibility & Uses \texttt{ConstructorInfo.params.length} and types; factory decision in host uses \texttt{interfaceToImpl}, not analyzer issues. \\
\hline
\end{tabular}
\end{table}

\subsection{Diagnostic mapping (host side, analyzer output)}

Although implemented in \texttt{extension.ts}, the mapping contract is part of the analyzer's consumer API:
\begin{itemize}
	\item constructor spans $\rightarrow$ Information diagnostic listing dependency names;
	\item \texttt{ConcreteTypeIssue} $\rightarrow$ Warning at parameter range;
	\item \texttt{CircularDependencyIssue} $\rightarrow$ Warning at constructor range;
	\item \texttt{MissingRegistrationIssue} $\rightarrow$ Warning at parameter range;
	\item parse guard \texttt{errors[0]} $\rightarrow$ Warning at \texttt{(0,0)} document start.
\end{itemize}

\section{Heuristics and design decisions}
\label{chap3:heuristics}

\subsection{Tree-sitter over Roslyn}

\begin{table}[htbp]
\centering
\caption{Roslyn vs.\ tree-sitter in this project}
\label{tab:roslyn-vs-ts}
\small
\begin{tabular}{|l|p{5.5cm}|p{5.5cm}|}
\hline
\textbf{Criterion} & \textbf{Roslyn} & \textbf{tree-sitter (chosen)} \\
\hline
Deployment in TS extension & Heavy (hosted process / bundled binaries) & Native addon + grammar \\
\hline
Constructor extraction & Exact & Exact for supported grammar \\
\hline
Cross-file symbols & Yes & No \\
\hline
Parse speed on edit & Moderate & Fast \\
\hline
Max buffer & Large & 32\,KiB in practice \\
\hline
\end{tabular}
\end{table}

\subsection{In-file class/interface sets}

Building sets from declarations in the same file avoids maintaining a solution-wide symbol table. Cost is false positives for concrete types when abstraction lives elsewhere---acceptable for a lint-style warning.

\subsection{Primitive and framework exclusion (file level)}

Only a small primitive list is excluded from missing-registration in the analyzer. Framework exclusions (\texttt{ILogger<T>}, etc.) are applied in the host because they require project-level conventions.

\subsection{Offset preservation}

Tree-sitter indices enable precise squiggles. The host converts to VS Code \texttt{Position} using document UTF-16 offsets. Mismatches can occur if the grammar byte offset model diverges from VS Code's UTF-16 indexing on rare surrogate pairs; samples use ASCII identifiers.

\section{Known limitations}
\label{chap3:limitations}

\begin{itemize}
	\item \textbf{32 KiB parse ceiling} --- large generated migrations are invisible to AST features; only host regex sees them.
	\item \textbf{No cross-file semantics} --- interface implementation in another file does not suppress concrete-type warnings in the consumer file.
	\item \textbf{Narrow registration regex in analyzer} --- \texttt{AddAutoMapper}, \texttt{AddDbContext}, factory delegates, and \texttt{TryAdd*} are not recognized file-locally.
	\item \textbf{Parameter type resolution} --- attributes, tuples, function pointers, and \texttt{global::} qualifiers may yield incomplete type strings.
	\item \textbf{Parse error nodes} --- \texttt{hasError} yields generic warning; no error-node recovery strategy.
	\item \textbf{No async/disposal analysis} --- \texttt{IAsyncDisposable} and \texttt{IDisposable} marker interfaces are not special-cased in the analyzer (host excludes markers from registration suggestions).
\end{itemize}

\section{Grammar node types referenced in code}
\label{chap3:node-types}

Table~\ref{tab:grammar-nodes} catalogs C\# grammar node \texttt{type} strings the analyzer matches. These names come from \texttt{tree-sitter-c-sharp} and appear in the Inspector or query output when debugging parse trees.

\begin{table}[htbp]
\centering
\caption{C\# grammar node types used in \texttt{analyzer.ts}}
\label{tab:grammar-nodes}
\small
\begin{tabular}{|l|p{10cm}|}
\hline
\textbf{Node type} & \textbf{Role} \\
\hline
\texttt{class\_declaration} & Extract class name; scope for constructors. \\
\hline
\texttt{interface\_declaration} & Build in-file interface set for concrete-type check. \\
\hline
\texttt{constructor\_declaration} & Extract parameter list span. \\
\hline
\texttt{parameter\_list} & Container for parameters. \\
\hline
\texttt{parameter}, \texttt{\_parameter\_array} & Single parameter; type/name fields. \\
\hline
\texttt{identifier}, \texttt{generic\_name} & Type and name text via \texttt{getText}. \\
\hline
\texttt{nullable\_type}, \texttt{predefined\_type}, \texttt{array\_type} & Alternate type shapes in \texttt{findTypeOfParameter}. \\
\hline
\end{tabular}
\end{table}

\section{Worked example: \texttt{CheckoutCoordinator}}
\label{chap3:worked-example}

Consider the following constructor (from \texttt{CheckoutSample/Services.cs}):

\begin{lstlisting}[caption={CheckoutCoordinator constructor (excerpt)}]
public CheckoutCoordinator(
    IPaymentGateway payment,
    IInventoryService inventory,
    IPricingEngine pricing,
    INotificationService notifications)
\end{lstlisting}

\textbf{Analyzer pass:} all parameter types are interfaces declared in other files in the same sample; no \texttt{ConcreteTypeIssue}. File-local missing-registration may fire if \texttt{CompositionRoot.cs} in the same workspace folder is not part of the same parse context---typically the analyzer analyzes one file at a time, so missing-registration in \texttt{Services.cs} flags parameters not registered \emph{in Services.cs itself}, not in sibling files.

\textbf{Host pass:} \texttt{buildWorkspaceSummary} merges registrations from \texttt{CompositionRoot.cs}, sees \texttt{IInventoryService} and \texttt{IPricingEngine} already registered, and lists \texttt{IPaymentGateway} and \texttt{INotificationService} as missing. Factory eligibility uses parameter count four $\geq 3$ and interface map entries for all types.

This example clarifies why both modules are necessary: the analyzer educates at file level; the host answers the composition question developers actually care about.

Chapter~\ref{chap5:evaluation} documents MealPlanner as the primary stress test for the parse guard and contrasts file-local vs.\ workspace missing-registration behavior on \texttt{ApplicationServiceExtensions.cs}.
