\selectlanguage{english}
\chapter[Extension host]{VS Code extension host and auxiliary modules}
\label{chap4:extension}
\begin{chapquote}{Microsoft Documentation}
``Extensions can interact with the editor through the VS Code Extension API.''
\end{chapquote}

\section{Purpose and responsibilities}
\label{chap4:purpose}

\texttt{extension.ts} is the \textbf{activation entry point} (\texttt{activate}/\texttt{deactivate}). It binds the analyzer to editor lifecycle events, exposes eleven contributed commands (see \texttt{package.json}), registers a \texttt{CodeActionProvider} for quick fixes, and coordinates file generation under \texttt{GeneratedDI/}.

Auxiliary modules isolate edit construction:
\begin{itemize}
	\item \texttt{registrationInsert.ts} --- find composition-root insertion points, deduplicate lines, normalize lifetimes;
	\item \texttt{extractInterface.ts} --- multi-file \texttt{WorkspaceEdit} for extract-interface quick fixes.
\end{itemize}

\section{Key data structures and interfaces}
\label{chap4:data-structures}

Beyond \texttt{WorkspaceDiSummary} (Chapter~\ref{chap2:data-structures}), the host uses:

\begin{table}[htbp]
\centering
\caption{Host-specific structures}
\label{tab:host-types}
\begin{tabular}{|l|p{8.8cm}|}
\hline
\textbf{Symbol} & \textbf{Role} \\
\hline
\texttt{WorkspaceCSharpFile} & Pair of \texttt{TextDocument} and source string. \\
\hline
\texttt{GeneratedFile} & \texttt{fileName} + generated C\# content for disk write. \\
\hline
\texttt{RegistrationInsertTarget} & URI, \texttt{insertPosition}, indentation, human label (\texttt{CompositionRoot}, etc.). \\
\hline
\texttt{MARKER\_SERVICE\_INTERFACES} & Exclude \texttt{IDisposable}, \texttt{ICloneable}, \ldots{} from interface$\rightarrow$impl suggestions. \\
\hline
\texttt{DiCodeActionProvider} & Implements \texttt{provideCodeActions} for DI diagnostics. \\
\hline
\end{tabular}
\end{table}

\section{Workspace aggregation algorithm}
\label{chap4:workspace-algo}

\texttt{buildWorkspaceSummary()} performs:
\begin{enumerate}
	\item Load up to 200 \texttt{.cs} files; skip unreadable paths.
	\item Per file: run \texttt{analyzeCSharp}; merge constructors; extend \texttt{registeredTypes} via \texttt{collectExtendedRegistrationTypes}.
	\item Detect \texttt{aspNetMvcHost} when \texttt{WebApplication.CreateBuilder} or \texttt{AddControllers*} appears.
	\item Build \texttt{interfaceToImpl} via \texttt{collectInterfaceImplementations} (regex on \texttt{public class Impl : IFace}).
	\item Mark \texttt{nonInjectableClassNames} for classes inheriting \texttt{Profile}, \texttt{Migration}, \texttt{Controller}, etc.
	\item Record namespaces and \texttt{CompositionRoot} presence.
\end{enumerate}

\texttt{collectExtendedRegistrationTypes} extends plain \texttt{Add*} scanning with:
\begin{itemize}
	\item \texttt{AddDbContext<T>} $\rightarrow$ \texttt{T}, \texttt{DbContextOptions<T>};
	\item \texttt{AddAutoMapper(...)} $\rightarrow$ \texttt{IMapper};
	\item \texttt{AddIdentity<TUser,TRole>} / \texttt{AddIdentityCore<TUser>} $\rightarrow$ Identity managers;
	\item \texttt{AddRoles}/\texttt{AddRoleManager} $\rightarrow$ \texttt{RoleManager<T>};
	\item \texttt{typeof(I)} , \texttt{typeof(Impl)} pairs in \texttt{ServiceDescriptor} style registrations.
\end{itemize}

\section{Extension commands and workflows}
\label{chap4:commands}

Table~\ref{tab:commands} maps user commands to workflows.

\begin{table}[htbp]
\centering
\caption{Contributed commands and workflows}
\label{tab:commands}
\small
\begin{tabular}{|p{4.1cm}|p{8.6cm}|}
\hline
\textbf{Command} & \textbf{Workflow} \\
\hline
\texttt{analyzeFile} & Resolve active/visible C\# document $\rightarrow$ \texttt{analyzeCSharp} $\rightarrow$ Output report + refresh diagnostics. \\
\hline
\texttt{suggestRegistrations} & File-scoped \texttt{buildRegistrationSuggestions} (interface pairs + concrete types). \\
\hline
\texttt{analyzeWorkspace} & \texttt{buildWorkspaceSummary} $\rightarrow$ list missing registrations (cross-file) and constructors by file. \\
\hline
\texttt{suggestWorkspaceRegistrations} & \texttt{buildWorkspaceRegistrationSuggestions} with lifetime from composition root. \\
\hline
\texttt{suggestFactories} & File-level factory snippets; fallback to workspace factory list if empty. \\
\hline
\texttt{suggestWorkspaceFactories} & \texttt{buildWorkspaceFactorySuggestions} (factory \texttt{.cs} only, no \texttt{AppBuilder} dump). \\
\hline
\texttt{generateWorkspaceFactories} & Write \texttt{GeneratedDI/*Factory.cs} + \texttt{AppBuilder.cs}. \\
\hline
\texttt{insertWorkspaceRegistrations} & Suggest lines $\rightarrow$ \texttt{insertRegistrationsIntoWorkspace}. \\
\hline
\texttt{mergeGeneratedIntoCompositionRoot} & Parse \texttt{GeneratedDI/AppBuilder.cs} $\rightarrow$ insert into composition root. \\
\hline
\end{tabular}
\end{table}

\textbf{Internal commands} (not in palette, invoked from code actions): \texttt{insertRegistrationForType}, \texttt{extractInterfaceAndRegister}.

\subsection{Registration suggestion logic}

\texttt{buildWorkspaceRegistrationSuggestions} emits:
\begin{itemize}
	\item \texttt{services.Add\{Lifetime\}<IInterface, Impl>()} for unregistered interface keys;
	\item \texttt{services.Add\{Lifetime\}<Concrete>()} for classes with constructors that are not factory targets, not already registered, and not excluded by MVC/test/non-injectable rules.
\end{itemize}

\texttt{preferredLifetimeForWorkspace} chooses \texttt{Scoped}, \texttt{Transient}, or \texttt{Singleton} from existing \texttt{services.Add*} usage in the composition root, else ASP.NET $\rightarrow$ Scoped, else Transient.

\subsection{Factory generation logic}

\texttt{shouldGenerateFactory} requires:
\[
|\text{parameters}| \geq 3 \quad \land \quad \forall p.\; \texttt{interfaceToImpl.has}(p.\text{type})
\]

\texttt{buildFactoryFileContent} emits a sealed class with \texttt{IServiceProvider} and \texttt{Create()} calling \texttt{GetRequiredService<T>()} for each dependency.

\texttt{buildAppBuilderFileContent} rules (refined):
\begin{itemize}
	\item Register \textbf{factory types only}, not the concrete service when a factory exists;
	\item Include interface$\rightarrow$implementation lines required by factory dependencies even if already registered elsewhere (standalone \texttt{AppBuilder} completeness);
	\item Skip marker interfaces and non-injectable classes.
\end{itemize}

\subsection{Registration insert module (\texttt{registrationInsert.ts})}

\texttt{findRegistrationInsertTarget} priority:
\begin{enumerate}
	\item File containing \texttt{class CompositionRoot};
	\item \texttt{public static IServiceCollection Add*Services(...)};
	\item \texttt{void ConfigureServices(IServiceCollection services)}.
\end{enumerate}

Insertion occurs at the \textbf{start of the line} containing \texttt{return services.BuildServiceProvider()} (or \texttt{return services;}), preserving indentation. \texttt{filterNewRegistrationLines} avoids duplicate substrings in normalized source.

\subsection{Extract interface module (\texttt{extractInterface.ts})}

Workflow for \texttt{extractInterfaceAndRegister}:
\begin{enumerate}
	\item Derive \texttt{I\{Concrete\}} name; locate implementation file.
	\item Create interface file in \texttt{Interfaces/} folder if present, else beside implementation.
	\item Append interface to class bases; replace constructor parameter type in consumer class.
	\item Insert \texttt{services.AddScoped<I, Concrete>()} via \texttt{insertRegistrationsIntoWorkspace}.
\end{enumerate}

\section{Diagnostics and quick fixes}
\label{chap4:diagnostics}

\texttt{analyzeCSharpDocument} maps analyzer output to \texttt{Diagnostic} entries (source \texttt{di-plugin}):
\begin{itemize}
	\item Information/hint on constructors with dependency lists;
	\item Warning on concrete types, cycles, file-local missing registrations;
	\item File-level parse limit warnings at range \texttt{(0,0)}.
\end{itemize}

\texttt{DiCodeActionProvider} offers:
\begin{itemize}
	\item local \texttt{ConfigureServices} insert when pattern matches;
	\item workspace insert command for missing types;
	\item register \texttt{I*} $\rightarrow$ concrete or full extract-interface workflow.
\end{itemize}

\section{Heuristics and design decisions}
\label{chap4:heuristics}

\begin{itemize}
	\item \textbf{Regex + AST hybrid:} AST for constructors; regex for registrations and inheritance (pragmatic for ASP.NET extension methods).
	\item \textbf{Marker interfaces excluded:} \texttt{IDisposable} on \texttt{UnitOfWork} must not produce \texttt{AddScoped<IDisposable, ...>}.
	\item \textbf{Non-injectable bases:} AutoMapper \texttt{Profile}, EF \texttt{Migration}, MVC controllers excluded from concrete registration suggestions.
	\item \textbf{Implicit framework services:} \texttt{IServiceProvider}, \texttt{ILogger<T>}, \texttt{IOptions<T>}, etc. excluded from workspace missing-registration reports.
	\item \textbf{Factory vs.\ registration separation:} reduces duplicate output between command categories (user-facing clarity).
\end{itemize}

\section{Known limitations}
\label{chap4:limitations}

\begin{itemize}
	\item \textbf{200-file workspace cap} may truncate very large solutions.
	\item \textbf{Interface detection} requires \texttt{I[A-Z]} prefix on base types; custom naming (\texttt{Abstractions.IEmail}) is missed.
	\item \textbf{Extract interface} creates empty interfaces; members are not copied from implementation.
	\item \textbf{Lifetime inference} is majority-vote on existing lines, not architecture-aware.
	\item \textbf{Merge/suggest} may still propose \texttt{JobScheduler}-style simple services while factories target complex ones---by design.
	\item \textbf{No undo grouping} for multi-file extract-interface edits beyond VS Code default \texttt{WorkspaceEdit} behavior.
	\item Commands \texttt{insertRegistrationForType} and \texttt{extractInterfaceAndRegister} are not listed in \texttt{package.json} contributes (palette) but are valid command IDs for code actions.
\end{itemize}
