\selectlanguage{english}
\chapter{Conclusions}
\label{conclusions}
\begin{chapquote}{Edsger W. Dijkstra}
``Simplicity is a great virtue but it requires hard work to achieve it and education to appreciate it.''
\end{chapquote}

\section{Summary of contributions}

This thesis presented the design, implementation, and manual evaluation of a Visual Studio Code extension that assists C\# developers using \texttt{Microsoft.Extensions.DependencyInjection}. The solution combines a tree-sitter--based analyzer module with a heuristic workspace aggregator hosted in the Extension API. Together they provide constructor dependency insight, DI smell detection, registration suggestions, optional factory code generation, and composition-root insertion workflows.

The modular split---\texttt{analyzer.ts} for syntax facts, \texttt{extension.ts} for orchestration, \texttt{registrationInsert.ts} and \texttt{extractInterface.ts} for targeted edits---demonstrates a practical architecture for small language-aware IDE tools that cannot assume Roslyn as a dependency. Each module's responsibilities, data structures, algorithms, commands, heuristics, and limitations were documented in dedicated chapters to support academic review and future maintenance.

\section{Evaluation against objectives}

Chapter~\ref{chap5:evaluation} answered the research questions posed in Chapter~\ref{chap1:research-questions}:

\begin{itemize}
	\item \textbf{RQ1 (constructor extraction):} tree-sitter reliably extracted constructors and parameter types on all sample files under the 32\,KiB limit; MealPlanner required skip-guard for EF snapshots without aborting the host.
	\item \textbf{RQ2 (workspace missing registrations):} extended regex registration detection plus implicit framework filters eliminated the majority of false positives on MealPlanner (\texttt{IMapper}, Identity managers, \texttt{IServiceProvider}).
	\item \textbf{RQ3 (factory generation):} the three-parameter rule with full interface mapping correctly included \texttt{OrderNotificationService}, \texttt{CheckoutCoordinator}, \texttt{ImportPipeline}, and \texttt{ExportPipeline} while excluding two-parameter services in Checkout and MultiFactory samples.
	\item \textbf{RQ4 (editor integration):} insert and merge into \texttt{CompositionRoot} succeeded after fixing return-line insertion; extract-interface quick fix validated on \texttt{ConcreteMailerService}.
\end{itemize}

Table~\ref{tab:conclusions-samples} recalls sample-level outcomes.

\begin{table}[htbp]
\centering
\caption{Sample validation summary (see Chapter~\ref{chap5:evaluation} for detail)}
\label{tab:conclusions-samples}
\begin{tabular}{|l|p{10cm}|}
\hline
\textbf{Sample} & \textbf{Key result} \\
\hline
\texttt{BaseSample} & Registration suggestions without existing composition root \\
\hline
\texttt{BuilderFactorySample} & Factory + concrete smell + merge \\
\hline
\texttt{CheckoutSample} & Factory positive / \texttt{PaymentCaptureService} negative \\
\hline
\texttt{MultiFactorySample} & Two factories in one workspace \\
\hline
\texttt{MealPlanner} & Scale + AutoMapper/Identity; parse guard essential \\
\hline
\texttt{EventPlanner} & \texttt{ServiceDescriptor} style registrations recognized \\
\hline
\end{tabular}
\end{table}

\section{Lessons learned during development}

Several regressions shaped the final architecture and are worth recording as engineering lessons:

\begin{enumerate}
	\item \textbf{Native binding limits are hard failures.} A single oversized file caused \texttt{EINVAL} and broke every command until per-file guards existed---fail-soft must be explicit, not assumed.
	\item \textbf{Registration detection must mirror real ASP.NET code.} File-local \texttt{AddScoped<T>} regex alone is insufficient for production APIs using \texttt{AddAutoMapper} and Identity.
	\item \textbf{User mental models separate factories from registrations.} Presenting identical \texttt{AppBuilder} output in both command categories created perceived bugs; separating outputs improved usability without changing core analysis.
	\item \textbf{Text insertion is fragile.} Token-level positions for \texttt{return} corrupted lines; line-start insertion is a robust anchor for composition roots.
\end{enumerate}

\section{Limitations}

The approach remains \textbf{heuristic and syntactic}, not a substitute for Roslyn-powered refactorings or runtime DI validation. Large C\# files exceed the tree-sitter buffer limit; cross-file concrete-type smells may persist; empty extracted interfaces require manual completion; lifetime inference does not encode domain rules about captive dependencies or singleton state.

Manual test procedures lack CI automation; external validity to arbitrary open-source repositories is unknown. The 200-file workspace cap may truncate analysis in very large solutions.

\section{Future work}

Recommended extensions include:
\begin{itemize}
	\item Roslyn integration or LSP collaboration for symbol-accurate analysis and partial-class merging;
	\item configurable factory thresholds and exclusion lists via \texttt{package.json} settings;
	\item golden-file Node tests executing \texttt{buildWorkspaceSummary} on each \texttt{sample/} folder;
	\item optional dependency graph export (Graphviz or Mermaid) for architecture reviews;
	\item member copying in extract-interface refactorings;
	\item regeneration of committed \texttt{GeneratedDI/} artifacts under CI to detect generator drift.
\end{itemize}

In summary, the DI Plugin shows that a focused VS Code extension can materially reduce friction in C\# dependency injection workflows when analysis rules are transparent, failures are contained per file, and validation is anchored in reproducible sample workspaces.
